Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-01-07 analysis/paper.qmd
Add introduction and expand discussion

---
title: Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 56 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on significantly smaller in the 
  Early Holocene compared to present conditions, with many species that are 
  found throughout the region's 'hilly flanks' today indicated to have much more
  restricted distributions centered on the Levant, Cyprus and Western Anatolia. 
  Ranges shrunk by on average c. 10% from the terminal Pleistocene to the Early 
  Holocene, with a further c. 10% reduction for the duration of the Younger 
  Dryas. However, modelled ranges do not reliably predict the observed 
  occurrence of specific species at archaeological sites. [...]
  The regional 
  ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Stop modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
    - [x] Comparison with archaeobot
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [ ] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
```

@@

* The first farming societies had an ecological context
* Subsistence is understood (largely) through archaeobotany and zooarchaeology; ecological context from environmental archaeology, palynology, palaeoclimate records, etc.
  * But these have a variety of biases (human selection, taphonomy, etc.)
  * And at the end of the day only represent specific places – interpolation to the entire region is not straightforward
  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. @MartinEtAl2016]
  * Or to fully contextualise subsistence strategies and shifts in strategies in response to environmental shifts [e.g. @YaworskyEtAl2023].
* Here we present an alternative approach using ENM
  * Whole region, at multiple climate snapshots
  * Independent of archaeobot. and pal. clim. data, so can verify and compare

@@

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Important to e.g.
  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2017; @MartinEtAl2025]
  * Reconstruct sequences of domestication [@YeomansEtAl2017]
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Geographically and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

@@

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworksyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik? --> [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].

@@

# Methods and materials

@@

```{r tbl-occ-count}
#| tbl-cap: "Recorded occurrence of flora considered in this study in Neolithic and contemporary West Asia"
flora |>
  mutate(
    exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis"),
    label = if_else(
      !is.na(taxon_detail),
      glue::glue("*{taxon}*\n\n<small>incl. {taxon_detail}</small>"),
      paste0("*", taxon, "*")
    )
  ) |>
  select(label, n_present, p_present, n_occ, exclude) |>
  arrange(-n_present) |>
  gt() |>
  cols_label(
    label = "Taxon",
    n_present = "Neolithic",
    n_occ = "Present"
  ) |>
  tab_spanner("Occurrences", columns = starts_with("n_")) |>
  cols_merge_n_pct(n_present, p_present) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  cols_align("left", c(label)) |>
  fmt_percent(p_present, decimals = 0) |>
  fmt_markdown(label) |>
  sub_missing() |>
  tab_options(footnotes.marks = "*")
```

@@

```{r fig-occ-map}
#| fig-cap: Georeferenced occurrence records from West Asia for the `r nrow(flora)` modelled taxa (N=`r nrow(unnest(flora, occ))`)
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

@@

```{r niche-predict}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = environment, .progress = TRUE)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
```

@@

## Model assessment

@@

```{r fig-model-sensitivity-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model sensitivity on test dataset
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "sens") |>
  mutate(label = if_else(.estimate < 0.85, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Sensitivity", labels = label_percent()) +
  theme_minimal_grid() 
```

@@

```{r niche-archaeo}
# TODO: factor out
# TODO: what about negatives?
archaeo_flora_predictions <- archaeo_flora |>
  filter(taxon %in% flora$taxon) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong) |>
  mutate(
    prediction = map(taxon, \(x) filter(flora, taxon == x)$prediction[[1]]),
    prediction = map2(prediction, geometry, function(pred, geom) {
      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
      st_extract(pred, sf_point)
    }),
    prediction = map2(prediction, climate_period, function(pred, per) {
      pred |>
        as_tibble() |>
        filter(as.character(period) == per) |>
        transmute(
          p_absent = prediction.FALSE.,
          p_present = prediction.TRUE.,
          pred_present = p_present > p_absent
        )
    })
  ) |>
  unnest(c(prediction))

@@

```{r niche-pred-range}
area_present <- function(pred, res = pred_res) {
  st_apply(pred, "period", sum, na.rm = TRUE)["pred_present"] |>
    data.frame() |>
    mutate(area_present = pred_present * (pred_res)^2) |>
    select(period, area_present)
}

flora <- flora |>
  mutate(
    # Binarize
    prediction = map(prediction, mutate, pred_present = prediction.TRUE. > 0.5),
    area_present = map(prediction, area_present)
  )
```

@@

Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene.
These include *Ficus carica* (fig);
*Hordeum* spp. (wild barleys);
*Lathyrus aphaca* and *L. sativus* (both marginally edible legumes);
*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see below);
*Aegilops speltoides*, but not *Aegilops tauschii* (goatgrasses);
and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
leaving the Levant and to a lesser extent Western Anatolia Cyprus as refugia.

Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @] and as a centre of origin of agriculture [@].
It should also be noted that the selection of flora for this study (@tbl-occ-count) is biased towards Levantine fauna due to the history of archaeological research in the region,
as is the occurrence dataset used to train the model [@fig-occ-map].
Within the Levant, retreat from the Badia.

Loss of the Northern Mesopotamia–Southeastern Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@KozlowskiAurenche2005], which puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication.
Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [@cite].
But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant)
and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
A scenario in which cultivation emerged at the edges of the ranges of valuable plant resources—as a means of extend their natural niche—is also plausible.

@@

Cyprus and W. Anatolia not considered part of the core zone – maybe they should be?

Exceptions to the trend of range reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable range centered on Northern Mesopotamia;
and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two very restricted ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
In the latter case, neither of these areas are part of the predicted modern distribution of wild emmer (centered around the Caucasus and Northern Mesopotamia), but it would be consistent with archaeological evidence for early cultivation at Tell Abu Hureyra and other sites in the Upper Euphrates.
On the other hand, the model for *Triticum turgidum dicoccum* has the lowest predictive performance of all our fitted models.

@@

Several crop progenitors reduced in range before cultivation?
Reduction in range from BA to EH very evident in Hordeum, Pistacia atlantica.

@@

@@

@@

<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->

# Conclusion

@@

niche_taxon <- "Aegilops tauschii"
niche_prediction <- filter(flora, taxon == niche_taxon)$prediction[[1]]

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 56 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on significantly smaller in the 
  Early Holocene compared to present conditions, with many species that are 
  found throughout the region's 'hilly flanks' today indicated to have much more
  restricted distributions centered on the Levant, Cyprus and Western Anatolia. 
  Ranges shrunk by on average c. 10% from the terminal Pleistocene to the Early 
  Holocene, with a further c. 10% reduction for the duration of the Younger 
  Dryas. However, modelled ranges do not reliably predict the observed 
  occurrence of specific species at archaeological sites. [...]
  The regional 
  ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [ ] Discussion
    - [x] General trends
    - [ ] Case studies
    - [x] Comparison with archaeobot
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-model-metrics
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [ ] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)

levant <- st_bbox(c(xmin = 33, xmax = 39, ymin = 30, ymax = 38))
s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
```

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
Even before the end of the Ice Age, these species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.

Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail [@cites].
Together with studies of charcoal [@cite], pollen [@cite], soil [@cite], and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
However, each of these sources of evidence is subject to the wide variety of taphonomic and recovery biases that are inherent in any direct record of the past.
They are also, by definition, records of the (human) environment at particular times and places.
Interpolating these snapshots to give a holistic picture of the regional ecologies is not straightforward – to date, it has tended to rely on non-explicit, inductive modelling.
The majority are also filtered through human action, producing a mixed single that makes it difficult to disentangle anthropic effects from the background of environmental change in this period of rapid climatic alteration.

In this paper we present a complementary, deductive approach based on ecological niche modelling.
Rather than inferring environmental conditions from preserved physical evidence, 
we predict the ranges of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
<!-- TODO which we can use to ... --->
Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.

@@

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Important to e.g.
  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2016; @MartinEtAl2025]
  * Reconstruct sequences of domestication [@YeomansEtAl2017]
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Geographically and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

@@

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik? --> [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].

@@

# Data and model

@@

```{r tbl-occ-count}
#| tbl-cap: "Recorded occurrence of flora considered in this study in Neolithic and contemporary West Asia"
flora |>
  mutate(
    exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis"),
    label = if_else(
      !is.na(taxon_detail),
      glue::glue("*{taxon}*\n\n<small>incl. {taxon_detail}</small>"),
      paste0("*", taxon, "*")
    )
  ) |>
  select(label, n_present, p_present, n_occ, exclude) |>
  arrange(-n_present) |>
  gt() |>
  cols_label(
    label = "Taxon",
    n_present = "Neolithic",
    p_present = "(%)",
    n_occ = "Present"
  ) |>
  tab_spanner("Occurrences", columns = c(n_present, p_present, n_occ)) |>
  # cols_merge_n_pct(n_present, p_present) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  cols_align("left", c(label)) |>
  cols_align("right", c(n_present, p_present, n_occ)) |>
  cols_width(label ~ pct(50), n_present ~ 10, p_present ~ 10, n_occ ~ 20) |>
  fmt_percent(p_present, decimals = 0) |>
  fmt_markdown(label) |>
  sub_missing() |>
  tab_options(
    footnotes.marks = "*",
    table.font.size = 11,
    table.width = pct(100),
    latex.use_longtable = TRUE
  )
```

@@

```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia for the ", nrow(flora), " modelled taxa (N=", nrow(unnest(flora, occ)))}
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

@@

# Model assessment

```{r niche-predict}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      pred = map(niche, collect_predictions),
      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = environment, .progress = TRUE),
      # Binarize
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
      palaeodist = map(palaeodist, mutate, absent = prediction.FALSE. > 0.5)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
```

@@

@@

```{r fig-model-sensitivity-vs-n_occ, warning=FALSE}
#| fig-cap: Model sensitivity by number of training occurrences
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "sens") |>
  mutate(label = if_else(.estimate < 0.85, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic", size = 10/.pt) +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Sensitivity", labels = label_percent()) +
  theme_minimal_grid() 
```

@@

```{r niche-archaeo}
# TODO: factor out
# TODO: what about negatives?
archaeo_flora_predictions <-

archaeo_flora |>
  filter(taxon %in% flora$taxon) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong) |>
  mutate(
    prediction = map(taxon, \(x) filter(flora, taxon == x)$palaeodist[[1]]),
    prediction = map2(prediction, geometry, function(pred, geom) {
      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
      st_extract(pred, sf_point)
    }),
    prediction = map(prediction, as_tibble),
    prediction = map2(prediction, climate_period, function(pred, per) {
      filter(pred, as.character(period) == per)
    }),
    prediction = map(prediction, select,
                     pred_present = present, pred_absent = absent)
  ) |>
  unnest(c(prediction))

@@

```{r niche-pred-range}
area_present <- function(pred, res = pred_res) {
  st_apply(pred, "period", sum, na.rm = TRUE)["present"] |>
    data.frame() |>
    mutate(area_present = present * (pred_res)^2) |>
    select(period, area_present)
}

flora <- mutate(flora, area_present = map(palaeodist, area_present))
```

@@

Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene.
These include *Ficus carica* (fig);
*Hordeum* spp. (wild barleys);
*Lathyrus aphaca* and *L. sativus* (both marginally edible legumes);
*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see below);
*Aegilops speltoides*, but not *Aegilops tauschii* (goatgrasses);
and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
leaving the Levant and to a lesser extent the Aegean and Cyprus as refugia.

Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @] and as a centre of origin of agriculture [@].
Within the Levant, retreat from the Badia.

Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyHopfWeiss2012; @KabukcuEtAl2021; @UlasEtAl2024].
Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant),
and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
A scenario in which cultivation emerged at the edges of the ranges of valuable plant resources—as a means of extend their natural niche—is also plausible.

@@

Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
Were these area part of the same broader 'interaction sphere' [@cite] that produce Neolithic agriculture in West Asia?

Exceptions to the dominant trend of range reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable range centered on Northern Mesopotamia;
and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two limited ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
In the latter case, neither of these areas are part of the predicted modern distribution of wild emmer (centered around the Caucasus and Northern Mesopotamia), but it would be consistent with archaeological evidence for early cultivation at sites in the Upper Euphrates [@Willcox2024].

@@

* Most cereal and legume crops predicted to be Levantine
* Cereals show diverging southern/northern ranges (legumes not so much)
* Not rye (Anatolia), chickpea (N. Mesopotamia)

```{r fig-hindcast-barely-levant, eval=FALSE, cache=TRUE}
#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
palaeodist_hordeum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
))

# TODO: expand period facet labels
ggplot() +
  facet_grid(taxon ~ period) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_x_continuous(breaks = c(34, 36, 38)) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(
    xlim = c(33, 39), ylim = c(30, 38),
    crs = w_asia_albers, default_crs = latlong
  ) +
  labs(fill = "P(present)", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

```{r fig-hindcast-wheat-levant, eval=FALSE, cache=TRUE}
#| fig-cap: Predicted palaeodistribution of wild wheat in the Levant
palaeodist_triticum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Triticum"))$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Triticum"))$taxon))
))

# TODO: expand period facet labels
ggplot() +
  facet_grid(taxon ~ period) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_triticum, prediction.TRUE.)) + 
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_x_continuous(breaks = c(34, 36, 38)) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(
    xlim = c(33, 39), ylim = c(30, 38),
    crs = w_asia_albers, default_crs = latlong
  ) +
  labs(fill = "P(present)", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

<!-- TODO binarized range figure other founder crops? -->

Wild barley [*Hordeum spontaneum*, @fig-hindcast-barely-levant] and pistachio [*Pistacia atlantica*, @fig] show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with them being brought into cultivation.
Plausible that this prompted cultivation?
These two species [@ArranzOtaeguiRoe2023] also see marked declines in the archaeobotanical record from the Early PPNA/Early PPNB (where they were amongst the most common taxa) to the Late PPNB and Late Neolithic.
However, this trend is not seen in most other crop progenitors.

Cereals / other progenitors (not all) reduced in range before cultivation?
Reduction in range from BA to EH very evident in Hordeum, less so in ...

<!-- TODO: composite wheat progentor figure -->

@@

<!-- TODO: figure with flax, Pistacia, Bolboschoenus -->

@@

<!-- TODO: figure with rye -->

@@

<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
Is inconsistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.

@@

niche_taxon <- "Lathyrus oleraceus"
niche_prediction <- filter(flora, taxon == niche_taxon)$palaeodist[[1]]
